---
title: "DynaSym_PhD_Workshop_2025"
output: pdf_document
date: "2025-05-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install Packages

```{r install packages}
#install.packages("odin")
#install.packages("bbmle")
#install.packages("pkgload")
#install.packages("devtools")
library("odin")
library("bbmle")
library("pkgload")
library("devtools")

```
## Setup The Model as Function

```{r model}
log_growth_model <- odin({
  
  deriv(n) <- r* n * (1 - n/K)    # the logistic growth equation
  
  initial(n) <- n_initial   # provide the initial densityto solver
  
  # below : transfer of parameter values from R
  n_initial <- user()
  r <- user()
  K <- user()
  
  
})
```

## Assign parameter values

```{r assign_log_model}

lgm_assigned_01 <- log_growth_model$new(n_initial = 0.01, r = 0.5, K = 1000000)
lgm_assigned_02 <- log_growth_model$new(n_initial = 0.01, r = 1, K = 1000000)
lgm_assigned_03 <- log_growth_model$new(n_initial = 0.01, r = 0.5, K = 2000000)
lgm_assigned_04 <- log_growth_model$new(n_initial = 0.1, r = 0.5, K = 1000000)


```

## Simulate the time series (Solve the equation)

```{r run_log_model}

lgm_output_01 <- lgm_assigned_01$run(seq(0,100, length = 1001))
lgm_output_02 <- lgm_assigned_02$run(seq(0,100, length = 1001))
lgm_output_03 <- lgm_assigned_03$run(seq(0,100, length = 1001))
lgm_output_04 <- lgm_assigned_04$run(seq(0,100, length = 1001))

```

```{r show_log_model_head}

#head(lgm_output_01)
head(lgm_output_04)

```
## Plot log growth

```{r plot_log_growth}

## plot 1st model plot
plot(lgm_output_01[,1],lgm_output_01[,2], type = "l")
#plot(lgm_output_04[,1],lgm_output_04[,2], type = "l")

## overlay other plots
lines(lgm_output_04[,1],lgm_output_04[,2], col = "blue" , lwd = 2)

```

## Lotka Voltera Comp model

```{r LV_comp_model}

LV_comp_model <- odin({
  
  deriv(n1) <- r1* n1 * (1 - (n1 + a12*n2)/K1)    #eqn for first competitior
  deriv(n2) <- r2* n2 * (1 - (a21*n1 + n2)/K2)    #eqn for second competitior
  
  
  initial(n1) <- n1_initial   # provide the initial density to solver
  initial(n2) <- n2_initial
  
  
  # below : transfer of parameter values from R
  n1_initial <- user()
  n2_initial <- user()
  r1 <- user()
  K1 <- user()
  r2 <- user()
  K2 <- user()
  a12 <- user()
  a21 <- user()
  
  
})
```

## Assign parameter values

```{r assign_LV_model}

##### first come first serve scenario
#LV_assigned_01 <-  LV_comp_model$new(n1_initial = 0.1,
#                                     n2_initial = 0.01,
#                                     r1 = 0.1, 
#                                     K1 = 10,
#                                     r2 = 0.1, 
#                                     K2 = 10,
#                                    a12 = 1,
#                                    a21 = 1)

##### Coexistence : K1 < K2/a21  and  K2 < K1/a12
LV_assigned_02 <-  LV_comp_model$new(n1_initial = 0.01,
                                     n2_initial = 0.01,
                                     r1 = 0.1, 
                                     K1 = 10,
                                     r2 = 0.1, 
                                     K2 = 10,
                                    a12 = 0.2,
                                    a21 = 0.3)

##### One species dominates : K1 > K2/a21 and K2 < K1/a12 (or vice versa)
LV_assigned_03 <-  LV_comp_model$new(n1_initial = 0.01,
                                     n2_initial = 0.01,
                                     r1 = 0.1, 
                                     K1 = 10,
                                     r2 = 0.1, 
                                     K2 = 10,
                                    a12 = 2,
                                    a21 = 0.9)

##### Competitive exclusion : K1 > K2/a21 and K2 > K1/a12 (depends on starting values)
LV_assigned_04 <-  LV_comp_model$new(n1_initial = 0.01,
                                     n2_initial = 0.01,
                                     r1 = 0.1, 
                                     K1 = 10,
                                     r2 = 0.1, 
                                     K2 = 10,
                                    a12 = 2,
                                    a21 = 3)


```

## Simulate the time series (Solve the equation)

```{r run_LV_comp_model}

#LV_output_01 <- LV_assigned_01$run(seq(0,100, length = 1001))
LV_output_03 <- LV_assigned_03$run(seq(0,200, length = 1001))


```


```{r show_LV_model_head}

#head(lgm_output_01)
head(LV_output_02)

```
## Plot LV growth

```{r plot_LV_growth}

## plot 1st speices and 2nd species plot
plot(LV_output_03[,1],LV_output_03[,2], type = "l", col = "red")


## overlay 2nd species plot
lines(LV_output_03[,1],LV_output_03[,3], type = "l", col = "blue")

```

## Rosenzweig Mac Arthur Model/ Paradox of Enrichment

```{r setup_RMA_model}

rma_model <- odin({
  
  ### the modified (beacause flexible exponent term q) Rosenzweig MacArthur consumer-resource model
  deriv(n) <- r* n * (1 - n/K)      -       fmax * n^(1+q)/ (nhalf^(1+q) + n^(1+q)) * p      #eqn for prey
  deriv(p) <-  fmax * n^(1+q)/ (nhalf^(1+q) + n^(1+q)) * p  -  m*p                           #eqn for predator
  
  
  initial(n) <- n_initial   # provide the initial density to solver
  initial(p) <- p_initial
  
  
  # below : transfer of parameter values from R
  n_initial <- user()
  p_initial <- user()
  r <- user()
  K <- user()
  fmax <- user()
  nhalf <- user()
  q <- user()
  m <- user()
  
  
})


```
## Assign Parameter values

```{r assign_params_rma_model}


rma_assigned_01 <-  rma_model$new(n_initial = 0.01,
                                     p_initial = 0.01,
                                     r = 1, 
                                     K = 1,
                                  fmax = 0.35 ,
                                 nhalf = 0.5,
                                     q = 0,  #creates type 2 functional response (q>0 = s shaped)
                                     m = 0.1   
)

```

## Run Model

```{r rma_run_model}

rma_output_01 <- rma_assigned_01$run(seq(0,1000,0.1))

```

```{r head_rma_model}

head(rma_output_01)


```
## Plot RMA growth

```{r plot_rma_growth}

## plot 1st speices and 2nd species plot
plot(rma_output_01[,1],rma_output_01[,3], type = "l", col = "blue")


## overlay 2nd species plot
lines(rma_output_01[,1],rma_output_01[,2], type = "l", col = "green")

```

## Creating a bifurcation diagram

```{r simulate_RMA_nhalf_gradient}

nhalf_input <- seq(0.25,0.75,0.001)

#library("foreach")
nhalf_gradient <- foreach(
  i = 1:length(nhalf_input),
  .combine = "rbind"   #combine various output using rbind
) %do% {
  
  ### assigning model parameter values
  rma_assigned_01 <-  rma_model$new(n_initial = 0.01,
                                     p_initial = 0.01,
                                     r = 1, 
                                     K = 1,
                                  fmax = 0.35 ,
                                 nhalf = nhalf_input[i],
                                     q = 0,  #creates type 2 functional response (q>0 = s shaped)
                                     m = 0.1   
)
  time_out <- seq(0,1000,0.1)
  
  #simulate the model 
  rma_output_01 <- rma_assigned_01$run(time_out)

  cut_time <- round(0.75*length(time_out))   
  
  rma_output_01 <- rma_output_01[cut_time:length(time_out),] #only taking the last quarter of the time series post burn out
  
  #only output min and max predator densities
  data.frame(
    min_prey = min(rma_output_01[,2]),
    max_prey = max(rma_output_01[,2]),
    min_pred = min(rma_output_01[,3]),
    max_pred = max(rma_output_01[,3])
  )
  
}


```


## Plot the bifurcation diagram

```{r plot_RMA_nhalf_gradient}

plot(
  x = c(nhalf_input, nhalf_input,nhalf_input,nhalf_input),
  y = c(nhalf_gradient$min_prey,nhalf_gradient$max_prey, nhalf_gradient$min_pred, nhalf_gradient$max_pred),
  type = "n",
  xlab = "nhalf",
  ylab = "population extremes"
)

points(nhalf_input,nhalf_gradient$min_prey, pch=16, cex=0.2,col="darkgreen")
points(nhalf_input,nhalf_gradient$max_prey, pch=16, cex=0.2,col="darkgreen")
points(nhalf_input,nhalf_gradient$min_pred, pch=16, cex=0.2,col="darkblue")
points(nhalf_input,nhalf_gradient$max_pred, pch=16, cex=0.2,col="darkblue")

```
## Choosing values for parameters

### Allometric scaling

Biological rates follow approx. a 0.75 scaling power law (I is any rate here)

$$I = I_0m^{0.75}$$
This scaling is **per capita**, meaning per individuals

However individuals differ across populations (e.g. one individual of algae is very different from one individual of daphnia)

One possible solution : 

finish notes from Bjorn's file, 

Now till lunch, we'll try to do a similar bifurcation analysis as yestrday. Now the x axis gonna be
body mass of the predator relative to the prey body mass

## The equation system in odin

```{r YImode_in_odin}

yi_model <- odin({
  
  deriv(n) <- n * (1-n) - y*x*n / (nhalf + n) * p
  deriv(p) <- e*y*x*n / (nhalf + n) * p - x * p 
  
  
  initial(n) <- n_initial   # provide the initial density to solver
  initial(p) <- p_initial
  
  
  # below : transfer of parameter values from R
  n_initial <- user()
  p_initial <- user()
  y <- user()
  x <- user()
  nhalf <- user()
  e <- user()
  
})

```
```{r assign_yim}

# x = x0*m^-0.25 # loss rate calc
#x0 = 0.314 (+/-)? for herbivores

yim_assigned <- yi_model$new(
        n_initial = 0.5,
        p_initial = 0.5,
        y = 10,
        x = 0.314,
        nhalf = 0.5,
        e = 0.45
)

```

```{r simulate_yim}

timesteps <- seq(0,5000,1)

yim_output_01 <- yim_assigned$run(timesteps)

```

```{r plot_yim}

plot(
  x = c(yim_output_01[,1],yim_output_01[,1]),
  y = c(yim_output_01[,2],yim_output_01[,3]),
  type = "n",
  xlab = "time",
  ylab = "population density"
)

lines(yim_output_01[,1], yim_output_01[,2], col="green") # resource, a plant
lines(yim_output_01[,1], yim_output_01[,3], col="blue") # predator, an invertebrate consumer


```

Here the consumer and the resource have the same size!

Example of a consumer which is 100 times larger ()

```{r assign_yim_100}

# x = x0*m^-0.25 # loss rate calc
#x0 = 0.314 (+/-)? for herbivores

yim_assigned_100 <- yi_model$new(
        n_initial = 0.5,
        p_initial = 0.5,
        y = 10,
        x = 0.314*100^(-0.25),
        nhalf = 0.5,
        e = 0.45
)

```

```{r simulate_yim_100}

timesteps <- seq(0,5000,1)

yim_output_100 <- yim_assigned_100$run(timesteps)

```

```{r plot_yim_100}

plot(
  x = c(yim_output_100[,1],yim_output_100[,1]),
  y = c(yim_output_100[,2],yim_output_100[,3]),
  type = "n",
  xlab = "time",
  ylab = "population density"
)

lines(yim_output_100[,1], yim_output_100[,2], col="green") # resource, a plant
lines(yim_output_100[,1], yim_output_100[,3], col="blue") # predator, an invertebrate consumer


```


## Bifuraction diagram

```{r simulate_yim_m_gradient}

log_m_input <- seq(0,5,length=2000)

library("foreach")
m_gradient <- foreach(
  i = 1:length(log_m_input),
  .combine = "rbind"   #combine various output using rbind
) %do% {
  
  ### assigning model parameter values
  
yim_assigned <- yi_model$new(
        n_initial = 0.5,
        p_initial = 0.5,
        y = 10,
        x = 10^(0.314 + -0.25*log_m_input[i]),
        nhalf = 0.5,
        e = 0.45
)


  time_out <- seq(0,5000,0.1)
  
  #simulate the model 
  yim_output <- yim_assigned$run(time_out)

  cut_time <- round(0.75*length(time_out))   
  
  yim_output <- yim_output[cut_time:length(time_out), ] #only taking the last quarter of the time series post burn out
  
  #only output min and max resource and consumer biomass
  data.frame(
    min_res = min(yim_output[,2]),
    max_res = max(yim_output[,2]),
    min_cons = min(yim_output[,3]),
    max_cons = max(yim_output[,3])
  )
  
}


```


## Plot the bifurcation diagram

```{r plot_yim_m_gradient}

plot(
  x = c(log_m_input, log_m_input,log_m_input,log_m_input),
  y = c(m_gradient$min_res,m_gradient$max_res, m_gradient$min_cons, m_gradient$max_cons),
  type = "n",
  xlab = "mass m",
  ylab = "biomass extremes"
)

points(log_m_input,m_gradient$min_res, pch=16, cex=0.2,col="darkgreen")
points(log_m_input,m_gradient$max_res, pch=16, cex=0.2,col="darkgreen")
points(log_m_input,m_gradient$min_cons, pch=16, cex=0.2,col="darkblue")
points(log_m_input,m_gradient$max_cons, pch=16, cex=0.2,col="darkblue")

```


## Nutrient uptake/Chemostat models

### Uptake model

**One nutrient**

$$
U = \frac{U_{max}N}{N_{half}+N}
$$

**two nutrients**

$$
U = min(\frac{U_{max1}N_1}{N_{half1}+N_1}, \frac{U_{max2}N_2}{N_{half2}+N_1})
$$
### The chemostat model

$$
\frac{dN_1}{dt} = D(S_1 - N_1) - min(\frac{U_{max1}N_1}{N_{half1}+N_1}, \frac{U_{max2}N_2}{N_{half2}+N_1})C
$$
$D$ turnover rate
$S$ supply rate
$N_{1,2}$ Nutrient concentrations
$C$ Consumer density (individuals, biovolume, biomass)


$$
\frac{dN_2}{dt} = D(S_2 - N_2) - min(\frac{U_{max1}N_1}{N_{half1}+N_1}, \frac{U_{max2}N_2}{N_{half2}+N_1})C
$$

$$
\frac{dC}{dt} = min(\frac{U_{max1}N_1}{N_{half1}+N_1}, \frac{U_{max2}N_2}{N_{half2}+N_1})C - D C
$$
## The equation system in odin

```{r chemostatmodel_in_odin}

chemostat_model <- odin({
  
  deriv(n1) <- d*(s1-n1) - min(umax1*n1/(nhalf1 + n1), umax2*n2/(nhalf2 + n2))*c
  deriv(n2) <- d*(s2-n2) - min(umax1*n1/(nhalf1 + n1), umax2*n2/(nhalf2 + n2))*c
  deriv(c) <- min(umax1*n1/(nhalf1 + n1), umax2*n2/(nhalf2 + n2))*c - d*c
  
  
  initial(n1) <- n1_initial # provide the initial density to solver
  initial(n2) <- n2_initial 
  initial(c) <- c_initial
  
  
  # below : transfer of parameter values from R
  n1_initial <- user()
  n2_initial <- user()
  c_initial <- user()
  d <- user()
  s1 <- user()
  s2<- user()
  umax1<- user()
  umax2<- user()
  nhalf1<- user()
  nhalf2<- user()
})

```


```{r assign_chemostat}

# x = x0*m^-0.25 # loss rate calc
#x0 = 0.314 (+/-)? for herbivores

chemostat_assigned <- chemostat_model$new(
  n1_initial = 1,
  n2_initial = 1,
  c_initial = 0.5,
  d = 0.05,
  s1 = 1.25,
  s2 = 0.5,
  umax1 = 2,
  umax2 = 0.5,
  nhalf1= 0.2,
  nhalf2 = 0.3
)

```

```{r simulate_chemostat}

timesteps <- seq(0,1000,0.1)

chemostat_output_01 <- chemostat_assigned$run(timesteps)

```

```{r show_chemostat_output}

head(chemostat_output_01)

```


```{r plot_chemostat}

plot(
  x = c(chemostat_output_01[,1],chemostat_output_01[,1],chemostat_output_01[,1]),
  y = c(chemostat_output_01[,2],chemostat_output_01[,3],chemostat_output_01[,4]),
  type = "n",
  xlab = "time",
  ylab = "population density"
)

lines(chemostat_output_01[,1], chemostat_output_01[,2], col="green", lwd = 3) # resource1, e.g. nitrogen
lines(chemostat_output_01[,1], chemostat_output_01[,3], col="darkgreen", lwd = 3) # resource1, e.g. phosphorous
lines(chemostat_output_01[,1], chemostat_output_01[,4], col="blue", lwd = 3) # consumer, e.g. algae


```

